---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

First, we change the local settings so that the English language will be used. This is necessary because of the names of days of the week, which we will need later:
```{r, echo=TRUE, results='hide'}
Sys.setlocale("LC_ALL", "English")
```

## Loading and preprocessing the data
First we unpack the zipped files and view the content:
```{r, echo=TRUE}
zipfile <- "./activity.zip"
files <- unzip(zipfile, exdir = ".")
files
```
The resulting file `r files` has the comma separated values and therefore we use the R command *read.csv* and see the structure and the first few lines of the data:
```{r, echo=TRUE}
data <- read.csv(files)
str(data)
head(data)
```
The column "date" in the resulting data frame is in the form of factor, so we transform it to the class Date and see the data structure again::
```{r, echo=TRUE}
data <- transform(data, date = as.Date(date, format = "%Y-%m-%d"))
str(data)
```
We check the summary of the data:
```{r, echo=TRUE}
summary(data)
```
If we look at the values of the column "interval", we find that they represent the hours and minutes within a day i.e. they are written in the form "hhmm".

## What is mean total number of steps taken per day?

First we calculate the total number of steps taken each day using the R function *aggregate* with the function *sum*:
```{r, echo=TRUE}
data.agg <- aggregate(steps ~ date, data = data, sum)
```
Then we present the summarized number of steps in the form of a histogram:
```{r, echo=TRUE}
hist(data.agg$steps, col="blue", main="Histogram of the total number of steps taken each day\n(with original missing values)", xlab="Steps", labels=TRUE, ylim = c(0, 40))
dev.copy(png, file = "./figures/Figure1.png")     ## Copy plot to a PNG file
dev.off()  ## Don't forget to close the PNG device!
```

and calculate the summary of the column steps:
```{r, echo=TRUE}
steps.sum <- summary(data.agg$steps)
steps.sum
```
The mean of total number of steps taken per day is therefore `r format(steps.sum["Mean"], scientific=FALSE)` and the median is `r format(steps.sum["Median"], scientific=FALSE)`.

## What is the average daily activity pattern?
We calculate the mean number of steps across all days for each 5-minute interval using the functions *aggregate* and *mean* and see the structure of the resulting data frame:   
```{r, echo=TRUE}
data.agg.5min <- aggregate(steps ~ interval, data = data, mean)
str(data.agg.5min)
```
Then we plot the time series:

```{r, echo=TRUE}
plot(data.agg.5min$interval, data.agg.5min$steps, type = "l", col="red", main="Time series plot of the average number of steps taken as\na function of the 5-minute time interval", xlab="Interval", ylab="Steps")
dev.copy(png, file = "./figures/Figure2.png")     ## Copy plot to a PNG file
dev.off()  ## Don't forget to close the PNG device!
```

We search for a 5-minute interval, which contains the maximum number of steps, as follows:
```{r, echo=TRUE}
interval.max <- data.agg.5min$interval[which.max(data.agg.5min$steps)]
```
The 5-minute interval `r interval.max` (i.e. at `r paste0(floor(interval.max/100),":",interval.max%%100, " AM")`) contains the maximum number of steps, that is, `r round(max(data.agg.5min$steps),0)` steps.

## Imputing missing values

We calculate the number of missing values in the dataset, as follows:
```{r, echo=TRUE}
missing <- sum(is.na(data$steps))
```

```{r echo=TRUE}
tab.miss <- table(data[is.na(data$steps),c(1,2)], useNA = "ifany")
tab.miss
```

The total number of missing values is `r missing`. We have `r length(tab.miss)` days where the data are completely missing. We choose a strategy to fill all the missing values in the dataset with the mean of each 5-minute interval. First we check if we have available the means for all intervals:
```{r, echo=TRUE}
miss.agg <- sum(is.na(data.agg.5min$steps))
```

The number of missing means is `r miss.agg`. All the means are therefore available. 

```{r, echo=TRUE}
data.filled <- merge(data,data.agg.5min, by = "interval", all = TRUE,
                     suffixes = c("",".mean"))
data.filled <- data.filled[ order(data.filled[,"date"], 
                                  data.filled[,"interval"]), ]
data.filled[is.na(data.filled$steps),"steps"] <- 
        data.filled[is.na(data.filled$steps),"steps.mean"]
```

Check missing values again:
```{r, echo=TRUE}
sum(is.na(data.filled$steps))
```
There are no missing values. We aggregate the data in a similar manner as above and present them  in the form of a histogram:
```{r, echo=TRUE}
data.filled.agg <- aggregate(steps ~ date, data = data.filled, sum)
hist(data.filled.agg$steps, col="green", main="Histogram of the total number of steps taken each day\n(with missing values filled)", xlab="Steps", labels=TRUE, ylim = c(0, 40))
dev.copy(png, file = "./figures/Figure3.png")     ## Copy plot to a PNG file
dev.off()  ## Don't forget to close the PNG device!
```

and calculate the their summary:
```{r, echo=TRUE}
steps.filled.sum <- summary(data.filled.agg$steps)
steps.filled.sum
```
The mean of total number of steps taken per day is now `r format(steps.filled.sum["Mean"], scientific=FALSE)` and the median is `r format(steps.filled.sum["Median"], scientific=FALSE)`. The mean is the same as in the first case. The median is now equal to the mean (`r format(steps.filled.sum["Median"], scientific=FALSE)`) comparing to the median in the first case(`r format(steps.sum["Median"], scientific=FALSE)`). In this case, imputing the missing values adds `r length(tab.miss)` days of missing data in the highest column of the histogram.  

## Are there differences in activity patterns between weekdays and weekends?
We create a new factor variable in the dataset with two levels – “weekday”
and “weekend” indicating whether a given date is a weekday or weekend
day:
```{r, echo=TRUE}
data.filled$typeday <- ifelse(weekdays(data.filled$date) %in% 
c("Saturday","Sunday"), "weekend", "weekday")
data.filled$typeday <- as.factor(data.filled$typeday)
```

We calculate the mean number of steps across all days for each 5-minute interval in the same way as in the first case, but we additionally group by weekday and weekend:    
```{r, echo=TRUE}
data.filled.agg.5min <- aggregate(steps ~ typeday + interval, data = data.filled, mean)
str(data.filled.agg.5min)
```
Then we plot the time series:

```{r echo=TRUE}
library(lattice)
xyplot(steps ~ interval | typeday, data.filled.agg.5min, type = "l", layout = c(1, 2),
       xlab = "Interval", ylab = "Number of steps")
dev.copy(png, file = "./figures/Figure4.png")     ## Copy plot to a PNG file
dev.off()  ## Don't forget to close the PNG device!
```

